---
title: "Metro gauges methods"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
```

Get the source code for this notebook on Github [here](https://github.com/khondula/river-cities).

```{r, message=FALSE}
library(tidyverse)
library(dataRetrieval)
library(tidycensus)
library(fs)
library(glue)
library(sf)
library(rosm)
library(ggrepel)
library(RStoolbox)
library(scales)
library(leaflet)
library(knitr)
```

Read in CBSA shapefile and filter to Metro areas (n = 392)

```{r}
data_dir <- "data"
cbsa_shp <- glue("{data_dir}/tiger/tl_2019_us_cbsa.shp")
cbsa_sf <- sf::st_read(cbsa_shp) %>% dplyr::filter(LSAD == "M1")
```

# Query NWIS for USGS gauges

Define function to find gages within bounding box of a metropolitan area polygon

```{r}
get_usgs_sites <- function(my_cbsa, savemap = TRUE){

  my_cbsa_sf <- cbsa_sf %>% dplyr::filter(GEOID == my_cbsa)
  my_bbox <- my_cbsa_sf %>% st_bbox()

  # look for usgs sites with discharge within bbox
  # use error handling to return null if no sites returned
  usgs_sites <- tryCatch({
    dataRetrieval::whatNWISsites(parameterCd = "00060", bBox = my_bbox)},
    error = function(e) {return(NULL)})
  # only proceed if there are gauges
  if(!is.null(usgs_sites)){
  # get info about sites
  usgs_sites_info <- readNWISsite(usgs_sites[["site_no"]]) %>%
    dplyr::select(site_no, state_cd, drain_area_va)
  # get info about record length
  site_data <- whatNWISdata(parameterCd = "00060", # discharge
                                 siteNumber = usgs_sites[["site_no"]]) %>%
    arrange(begin_date) %>%
    mutate(early_enough = begin_date < "1950-01-01") %>%
    mutate(late_enough = end_date > "2019-01-01") %>%
    mutate(count_10obs = count_nu > 10)
  
  site_data_goodrecords <- site_data %>% 
    dplyr::filter(early_enough, late_enough, count_10obs)
  
  if(nrow(site_data_goodrecords) > 0){
    
  good_sites_sf <- site_data_goodrecords %>% 
    dplyr::select(site_no, dec_lat_va, dec_long_va) %>% 
    distinct() %>%
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"), crs = 4326)

  # make a table with site info of interest and add city name and pop
  df <- site_data_goodrecords %>%
    dplyr::select(agency_cd, site_no, station_nm, dec_lat_va, dec_long_va,
                  huc_cd, data_type_cd, parm_cd, stat_cd, begin_date, end_date,
                  count_nu) %>%
    mutate(site_no = as.character(site_no)) %>%
    arrange(site_no) %>%
    left_join(usgs_sites_info) %>%
    mutate(cbsa_geoid = my_cbsa_sf$GEOID,
           cbsa_name = my_cbsa_sf$NAME)
  
  gauges_dir <- glue("{data_dir}/gauges-cbsa-M1")
  if(!fs::dir_exists(gauges_dir)){fs::dir_create(gauges_dir)}
  if(nrow(df) > 0) {readr::write_csv(df, path = glue("{gauges_dir}/cbsa-{my_cbsa_sf$GEOID}.csv"))}
  
  if(savemap){
    myextent <- sp::bbox(as(my_cbsa_sf, "Spatial"))
    osm1 <- osm.raster(myextent)
    my_cbsa_prj <- sf::st_transform(my_cbsa_sf, 3857)
    good_sites_prj <- sf::st_transform(good_sites_sf, 3857)
    good_sites_prj$lat <- st_coordinates(good_sites_prj) %>% 
      as.data.frame() %>% pull(Y)
    good_sites_prj$lon <- st_coordinates(good_sites_prj) %>% 
      as.data.frame() %>% pull(X)

    m1 <- ggRGB(osm1, ggObj = TRUE, r = 1, g = 2, b = 3)
    m2 <- m1 +
      geom_sf(data = my_cbsa_prj, 
              color = "purple",
              fill = "purple", 
              alpha = 0.05) +
      geom_sf(data = good_sites_sf) +
      ggrepel::geom_text_repel(
        data = good_sites_prj,
        aes(x = lon, y = lat, label = site_no),
        nudge_y = 0.1
      ) +
      theme_minimal() +
      ggtitle(glue::glue("{my_cbsa_sf$NAME}")) +
      xlab(element_blank()) +
      ylab(element_blank())
   
    map_dir <- glue::glue("{data_dir}/maps-gauges-M1/")
    if(!fs::dir_exists(map_dir)){fs::dir_create(map_dir)}
      mapfile <- glue::glue("{map_dir}/gauges-M1-1950_{my_cbsa_sf$NAME}.pdf")
      pdf(mapfile, height = 10, width = 8)
      print(m2)
      dev.off()
  }
    }
  }
}

```

Use function for all 392 metropolitan areas
```{r, cache=TRUE, message=FALSE}
purrr::walk(cbsa_sf[["GEOID"]], ~get_usgs_sites(.x, savemap = FALSE))
```

Then read in table for each city and save as one table

```{r}
gauges_dir <- glue("{data_dir}/gauges-cbsa-M1")
df <- fs::dir_ls(gauges_dir) %>% 
  purrr::map_df(~read_csv(.x, col_types = c("cccddccccDDdcdcc")))

df %>% readr::write_csv(glue("{data_dir}/gauges-cbsa-M1.csv"))
```

View first few rows of data: 

```{r, echo = FALSE, results = 'as.is'}
knitr::kable(head(df[,c(1:3,10:16)]))
```

# Function to map gauges

Define function to see gauge locations on interactive map

```{r}
leaflet_gauges_map <- function(my_cbsa_geoid){
  gauges_df <- readr::read_csv(glue("{data_dir}/gauges-cbsa-M1.csv"),
                               col_types = "cccddccccDDdcdcc")
  gauges_geoids <- gauges_df[["cbsa_geoid"]] %>% unique()
  if(!my_cbsa_geoid %in% gauges_geoids){stop("No good gauges")}
  
  good_sites_sf <- gauges_df %>%
    dplyr::filter(cbsa_geoid == my_cbsa_geoid) %>%
    dplyr::select(site_no, station_nm, dec_lat_va, dec_long_va, 
                  begin_date, end_date, count_nu, drain_area_va) %>%
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"), crs = 4269)
  
  # metro areas
  cbsa_shp <- glue("{data_dir}/tiger/tl_2019_us_cbsa.shp")
  cbsa_sf <- sf::st_read(cbsa_shp) %>% dplyr::filter(LSAD == "M1")
  my_cbsa_sf <- cbsa_sf %>% dplyr::filter(GEOID == my_cbsa_geoid)
  my_bbox <- my_cbsa_sf %>% st_bbox()
  
  nhd <- "https://hydro.nationalmap.gov/arcgis/services/nhd/MapServer/WMSServer"
  wbd <- "https://hydro.nationalmap.gov/arcgis/services/wbd/MapServer/WMSServer"
  
  ll <- leaflet() %>%
    addProviderTiles(providers$Esri.WorldImagery, 
                     group = "Satellite Imagery") %>%
    addTiles(group = "Open Street Map") %>%
    addPolygons(data = my_cbsa_sf, popup = ~NAMELSAD,
                fillOpacity = 0) %>%
    addMarkers(data = good_sites_sf, 
               popup = ~glue("{site_no}<br>{station_nm}<br> 
               {begin_date} to {end_date}<br>n={count_nu}<br>
                            {drain_area_va} sq mi")) %>%
    addWMSTiles(nhd, 
                layers = list("5", "6", "9", "10"),
                options = WMSTileOptions(format = "image/png",
                                         transparent = TRUE),
                group = "NHD") %>%
    addWMSTiles(wbd, layers = list("7"),
                options = WMSTileOptions(format = "image/png",
                                         transparent = TRUE),
                group = "watersheds") %>%
    addLayersControl(baseGroups = c("Open Street Map", "Satellite Imagery"),
      overlayGroups = c("NHD", "watersheds")) %>%
    leaflet::hideGroup(c("watersheds"))
  
  return(ll)
}

```

To view gauges for a given city based on its [GEOID](https://github.com/khondula/river-cities/blob/master/data/reference-tables/cbsa-geoids.csv). Click a marker to see info about the site and discharge record. 

```{r, warning=FALSE}
leaflet_gauges_map("10500")
```


